\Large Last Writer Slices and Communication Traps

### 2013-11-02 paper.tex
committing a draft with more figure edits and some other fun changes

Data dependence information is directly useful during program debugging.  Prior
work has proposed a wide variety of dynamic program slicing
techniques~\cite{thinslicing,tipslicingsurvey}.  These techniques record a
subset (slice) of dynamic program instructions selected using data and control
dependence information.  Dynamic dependence slices expose the provenance of
data in the execution.  Work on debugging~\cite{whylinechi,whylineicse} in the
HCI and Software Engineering community shows that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.

@@

Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the program
crashes when Thread 2 tries to acquire the lock before Thread 1 has initialized
it.  Assuming the run time environment supports it, a core dump delivered to
the programmer after such a failure reveals very little about how to fix the
bug.  The lock is uninitizalized, so it may hold {\tt NULL} or it may hold
random bits.  Both scenarios provide no insight into the root cause of the
failure, which is that the code permits a use of the lock before its
initialization.   Comparing the core dump from the failing run to the memory
image from a correct execution -- a likely debugging strategy -- reveals only
that in a correct execution, the lock contains valid mutex state.  This
revalation is of limited use as it does not rule out arbitrary data corruption
as the bug's root cause.

The vital, missing piece to solving this bug is provenance information.
Debugging this bug would be easier if the core dump included information about
the thread and code point that last wrote each variable.  The programmer would
see that no thread had initialized the lock in the failing execution, but that
Thread 1 initialized the lock in non-failing executions.  This information
rules out data corruption as a root cause and guides the programmer to add
synchronization ensuring the correct ordering relationship.

Another facet to debugging this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs an
instrumented or tooled version of the program, the failure may not manifest in
a reasonable amount of time and running it may require unavailable real-world
inputs.  Thus, collecting provenance or other information offline is
time-consuming or impossible.  A better strategy is to continuously collect
that information in deployed systems and package it with core dumps sent to
programmers with bug reports.  The last writer slicing design we propose in
this work espouses this {\em in situ} data collection strategy and does so with
overheads appropriate for production systems.

@@

A good first example of a system that monitors inter-thread dependences is
software transactional memory (STM)~\cite{stm}.  In STMs, dependences
correspond to interference between accesses in different threads that must be
detected to rollback conflicting transactions.  Deterministic concurrent
runtime systems~\cite{coredet,grace} are similar -- they monitor
inter-thread dependences, to determine when to apply deterministic ordering
constraints on threads' accesses.  STMs and deterministic systems must track
all dependences and they must track them precisely, to provide strong
guarantees as programming and execution models.  Unfortunately, precision can
have a high runtime overhead -- {\em e.g.}, CoreDet incurs a mean overhead of
around 5x~\cite{coredet}.  This overhead is too high for most production
systems.

@@

In this work, we select a subset of inter-thread dependences in a systematic
way, rather than choosing an {\em ad hoc} subset.  \ctraps tracks dependences
that correspond to inter-thread communication.  \ctraps provides enough
precision to build useful analyses and runtime systems with overhead low enough
for use in production.

\section{Last Writer Slices}
\label{sec:lastwriterslices}

A last writer slice is a record associated with a memory location that stores
the thread and program point that last wrote data to that memory location.  A
last writer slice is a dynamic property of a memory location at a point in a
program's execution.   We formalize last writer slices in
Section~\ref{sec:lwssoundness}.

We first discuss last writer slices in more depth by describing the design of
our support for collecting them.  The system support consists of a runtime
library and a compiler analysis.  In Section~\ref{sec:debugging} we highlight
the role of last writer slices in debugging and show they help diagnose real
bugs, using case studies.

\subsection{Runtime Support: Maintaining the Last Writer Table}
We gather last writer slices at runtime by maintaining a {\em last writer
table} (\lwt).  The \lwt is a {\em shadow memory} that holds one entry for each
memory address accessed by the program.  Each \lwt entry records the thread and
program point that last wrote to the entry's corresponding memory location.
When a thread performs a write to a memory location, it updates the \lwt by
recording its thread identifier and the program point of its write operation to
the memory location's \lwt entry.  A program point may be an
instruction address, or a call stack.  We use instructions by default, but our
system supports 2-address bounded call-stacks as well.

Figure~\ref{fig:basicLWT} shows the basic operation of our runtime support.
The right side of the figure shows a program execution with three threads.  On
the left is the \lwt at the point of each access.  There are two key things to
notice.  First, write operations update the \lwt entry of the memory location
they access.  Second, there is no action necessary for read operations.  Making
read operations cheap is key to making the overhead of last writer slicing low
enough for deployment use.

@@

\subsection{Compiler Support: Analysis and Instrumentation}
Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call into our runtime
system just before each write to a memory location that may be accessed by code
executing in another thread (escaping memory accesses).  Write instrumentation
calls are passed three pieces of information: the address of the memory word
involved in the access, the thread identifier of the accessing thread, and the
accessing program point.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
The WhyLine debugger~\cite{whylineicse} demonstrated that it is useful to
enable programmers to make data provenance queries during debugging.  To answer
such questions, prior work has mostly used dependence slices that include
intra-thread dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The
overhead of slicing dependence chains and tracking temporal sequences is often
high, relegating such techniques to working on traces~\cite{whylinechi} or
debugging executions run ``in the lab''.

The \lwt dynamically records information about write operations -- what thread
and code point last wrote each memory location.   These last writer slices do
not explicitly encode anything about read operations.  However, by looking at
the \lwt at a point in the execution, the provenance of the data being read or
over-written can be determined.  For example, if an assertion failure occurs
due to an erroneous inter-thread data flow, the provenance of the data in the
assertion's condition helps explain why the failure occurred.  Last writer
slices therefore answer an important class of ``why?'' questions related to
concurrency errors.

Computing last writer slices is fast enough to deploy and use continuously (see
Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.  Thus, last
writer slices can be made part of the standard debugging information provided
when a program fails.

\subsubsection{Case Studies}

We illustrate the debugging benefits of last writer slices using a set of case
studies.  In each case, we debug a buggy program using last writer slices.  We
show that for some concurrency errors, last writer slices are a good match
because they provide answers to important {\em ``why?''}
questions~\cite{whylineicse}.

\paragraph{Case Study: HTTrack}

@@

\section{Evaluation}
\label{sec:eval}
We evaluated our design along several axes. Recall that in
Section~\ref{sec:debugging} we illustrated the debugging benefits of last
writer slices using case studies.   We evaluate our performance overheads to
demonstrate that last writer slices and \ctraps are feasible for use in
deployed systems.  We discuss the sources of \ctraps' overhead.   We show that
\ctraps is useful by implementing two existing applications and analyzing their
performance.  We characterize our instrumentation and the precision and
performance impact of our redundancy analysis.

>>>>

Data provenance information is directly useful during program debugging.
Work on debugging~\cite{whylinechi,whylineicse} in the HCI and Software
Engineering community shows that when programmers debug programs, they benefit
from the ability to directly ask such provenance (``why'') questions of their
debugger.

@@

Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the
program crashes when Thread 2 tries to acquire the lock before Thread 1 has
initialized it.  Assuming the run time environment supports it, a core dump
delivered to the programmer after such a failure reveals very little about how
to fix the bug.  The lock is uninitizalized, so the core dump may show that it
holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
insight into the root cause of the failure, which is that the code permits a
use of the lock before its initialization.   Comparing the core dump from the
failing run to the memory image from a correct execution -- a helpful debugging
strategy -- reveals only that in a correct execution, the lock contains valid
mutex state.  This revalation is of limited use as it does not rule out
arbitrary data corruption as the bug's root cause.

The vital, missing piece to solving this bug is provenance information.
Debugging would be easier if the core dump included information about the
thread and code point that last wrote the lock.  The programmer would see that
no thread had initialized the lock in the failing execution, but that Thread 1
initialized the lock in non-failing executions.  This information guides the
programmer to fix the bug by adding synchronization that ensures the correct
operation ordering.

Another challenge in debugging this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs the
program in a debugger, the failure may not manifest in a reasonable amount of
time and running it may require unavailable real-world inputs.  Thus,
collecting provenance or other information offline is time-consuming or
impossible.  A better strategy is to continuously collect that information in
deployed systems and package it with core dumps sent to programmers with bug
reports.  In this work, we propose a last writer slicing design that espouses
this {\em in situ} data collection strategy and does so with overheads
appropriate for production systems.

@@

Monitoring inter-thread dependences is the foundation of many useful
concurrency analyses.  For example, in a software transactional memory (STM)
system~\cite{stm} dependences correspond to interference between accesses in
different threads that must be detected to rollback conflicting transactions.
Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
they monitor inter-thread dependences, to determine when to apply
deterministic ordering constraints on threads' accesses.  STMs and
deterministic systems must track all dependences and they must track them
precisely, to provide strong guarantees as programming and execution models.
Unfortunately, precision has a high runtime overhead -- {\em e.g.},
CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
too high for use in production.

@@

The design for \ctraps that we develop in this work monitors a subset of
inter-thread dependences that correspond to inter-thread communication.
\ctraps provides enough precision to build useful analyses and runtime systems
with overhead low enough for use in production.

\section{Last Writer Slices}
\label{sec:lastwriterslices}
A last writer slice is a dynamic property of a memory location at a point in a
program's execution.   A memory location's last writer slice is a record
containing the program point and the identifier of the thread that last wrote
data to that memory location.  We gather last writer slices at runtime by
maintaining a {\em last writer table} (\lwt) that maps from a memory
location's address to its last writer slice.  Just before a thread writes to a
memory location,  the thread first finds the memory location's entry in the
\lwt and updates that entry with its thread identifier and its current point in the
program.  After updating the \lwt, the thread performs the original memory
access.

Figure~\ref{fig:basicLWT} shows the basic operation of the \lwt.  The right
side of the figure shows a program execution with three threads.  On the left
is the \lwt at the point of each memory access.  There are two key things to
notice.  First, write operations update the \lwt entry of the memory location
they access.  Second, there is no action necessary for read operations.
Making read operations cheap is key to making the overhead of last writer
slicing low enough for deployment use.

@@

\subsection{Debugging with Last Writer Slices}
\label{sec:debugging}
%To answer data provenance questions during debugging, prior work has mostly
%used dependence slices that include intra-thread
%dependences~\cite{whylineicse,conseq,tipslicingsurvey}.  The overhead of
%slicing dependence chains and tracking temporal sequences is often high,
%relegating such techniques to working on traces~\cite{whylinechi} or debugging
%executions run ``in the lab''.

At any point in an execution, the information stored in a memory location's
\lwt entry represents the provenance of the data stored in that memory
location.   Last writer slices do not explicitly record any information about
read operations and they do not need to to be useful.  Interrupting the
execution while debugging and examining an entry in the \lwt reveals the
provenance of a value just when it is needed.  For example, if an assertion
failure occurs due to an erroneous inter-thread data flow, the provenance of
the data in the assertion's condition helps explain why the failure occurred.
Computing last writer slices is fast enough to deploy and use continuously
(see Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.
Thus, last writer slices can be made part of the standard debugging
information provided when a program fails.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
ordering violation in HTTrack.} The provenance information stored in the last
writer slice for the mutex variable helps understand the bug's root cause.}
\end{figure}

%The core dump does not provide any
%insight into why the mutex is invalid at the time of the call or what its
%value should have been instead.

Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
that the the program crashes at point {\bf B} when it calls {\tt
hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
The last writer slices in the figure make this root cause clear: in the
failing execution, the last writer slice indicates the mutex is uninitialized; in
the non-failing execution, the last writer slice indicates the mutex was 
initialized by Thread 1 at point {\bf A}.  This information tells the
programmer that there is an ordering relationship between {\bf A} and {\bf B}
that must be enforced to fix the bug.  Note that without last writer slices,
the program crashes at {\bf B} and the core dump provides no information about
the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.

Another important trait of last writer slices illustrated by
Figure~\ref{fig:httlws} is that last writer slices are reported back to the
programmer {\em with the core dump on a failure}.  Capturing good debugging
information is especially important for concurrency bugs that manifest very
rarely or in the presence of only peculiar real-world inputs.  A primary goal
of this work is to show that we can collect last writer slices with overheads
that are low enough for production use so they can be included pervasively in
bug reports.

%We discuss last writer slices in more depth by first describing our run time
%support for collecting them.  The system support consists of a runtime library
%and a compiler analysis.  Section~\ref{sec:debugging} shows how last writer
%slices help with debugging using our running example from
%Figure~\ref{fig:bgcoredumpfail}.  We formalize last writer slices in
%Section~\ref{sec:lwssoundness}.

\subsection{Runtime \& Compiler Support}
 The \lwt is implemented as a {\em shadow memory} that holds one entry for
each memory address accessed by the program.  Each \lwt entry records the
thread and program point that last wrote to the entry's corresponding memory
location.  When a thread performs a write to a memory location, it updates the
\lwt by recording its thread identifier and the program point of its write
operation to the memory location's \lwt entry.  A program point may be an
instruction address, or a call stack.  We use instructions by default, but we experimented with 2-address bounded call-stacks as well.

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call into our runtime
system just before each write to a memory location that may be accessed by code
executing in another thread (escaping memory accesses).  Write instrumentation
calls are passed three pieces of information: the address of the memory word
involved in the access, the thread identifier of the accessing thread, and the
accessing program point.

@@

@@

\section{Evaluation}
\label{sec:eval}
We evaluated our design along several axes.  We illustrate the debugging
benefits of last writer slices in Section~\ref{sec:eval:debugging} using case
studies conducted using real, buggy programs on real.   We evaluate our
performance overheads to demonstrate that last writer slices and \ctraps are
feasible for use in deployed systems.  We discuss the sources of \ctraps'
overhead.   We show that \ctraps is useful by implementing two existing
applications and analyzing their performance.  We characterize our
instrumentation and the precision and performance impact of our redundancy
analysis.

\subsection{Debugging with Last Writer Slices}
\label{sec:eval:debugging}
We illustrate the debugging benefits of last writer slices using three case
studies.

\paragraph{Case Study: HTTrack}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:httdebug} describes the benefit provided by
last writer slices in debugging an ordering violation in HTTrack-3.43.9.
